---
title: "Urban Performance Scenario Analysis"
subtitle: "Technical Brief & Strategic Assessment"
author: "Urban Performance Modeling Team"
date: today
format:
  pdf:
    pdf-engine: pdflatex
    keep-tex: true
    documentclass: article
    geometry: "top=2cm, bottom=2.5cm, left=2cm, right=2cm"
    colorlinks: true
    toc: false
params:
  project_name: "Cancún Pok Ta Pok Assessment"
  base_year: 2025
  horizon_year: 2030
  matrix_csv: ""
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{xcolor}
  - \definecolor{mightyslate}{HTML}{556270}
  - \definecolor{pacifica}{HTML}{4ECDC4}
  - \definecolor{applechic}{HTML}{C7F464}
  - \definecolor{cheerypink}{HTML}{FF6B6B}
  - \definecolor{grandma}{HTML}{C44D58}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  dev = "pdf",
  fig.align = "center"
)

library(ggplot2)
library(dplyr)

# Load data matrix
matrix_df <- if (nzchar(params$matrix_csv) && file.exists(params$matrix_csv)) {
  read.csv(params$matrix_csv, stringsAsFactors = FALSE)
} else {
  data.frame(
    No = 1:27,
    Indicator = c(
      "Total Population", "Urban Footprint", "Population Density", "Land Consumption", "Infill Development",
      "Biodiversity Land Consumption", "Agricultural Land Consumption", "Green Land Consumption", "Green Land Per Capita", "Land Cover Loss",
      "Roads Length", "Roads Density", "Intersection Density", "Cycle Track Density", "Cycle Proximity",
      "Public Transport Proximity", "Amenities Proximity", "Jobs Proximity", "Hazard Exposure", "Tree Canopy & Carbon",
      "Housing Costs", "Anthropogenic Hazard", "Vacant Housing Rate", "Walkability Index", "Housing Affordability Ratio",
      "Urban Heat Island Intensity", "Transit Utilization Rate"
    ),
    Category = c(rep("Demographics & Land Use", 5), rep("Land Consumption & Ecology", 5), rep("Mobility & Infrastructure", 4), rep("Accessibility & Exposure", 5), rep("Risk & Resilience", 3), rep("Housing & Equity", 5)),
    Baseline = c(911500, 142.5, 6396, 0, 0, 0, 0, 0, 12.4, 0, 1250.4, 8.77, 48.2, 0.85, 24.5, 64.2, 52.8, 68.1, 18.4, 3450, 780, 12.3, 9.8, 46.2, 4.8, 3.2, 54.0),
    Horizon = c(1065000, 158.2, 6732, 15.7, 6.2, 0.8, 3.4, 1.1, 14.8, 15.7, 1380.0, 8.72, 52.1, 1.45, 41.2, 78.5, 66.4, 74.3, 14.1, 4820, 820, 9.5, 7.2, 58.7, 4.2, 2.8, 68.5),
    Abs_Change = c(153500, 15.7, 336, 15.7, 6.2, 0.8, 3.4, 1.1, 2.4, 15.7, 129.6, -0.05, 3.9, 0.6, 16.7, 14.3, 13.6, 6.2, -4.3, 1370, 40, -2.8, -2.6, 12.5, -0.6, -0.4, 14.5),
    Pct_Change = c(16.8, 11.0, 5.3, 100, 100, 100, 100, 100, 19.4, 100, 10.4, -0.6, 8.1, 70.6, 68.2, 22.3, 25.8, 9.1, -23.4, 39.7, 5.1, -22.8, -26.5, 27.1, -12.5, -12.5, 26.9),
    Units = c("inhabitants", "km²", "inhab/km²", "km²", "km²", "km²", "km²", "km²", "m²/capita", "km²", "km", "km/km²", "intersections/km²", "km/km²", "% pop", "% pop", "% pop", "% pop", "% pop", "t CO2/yr", "USD/m²", "% pop", "% units", "index [0-100]", "PIR ratio", "°C delta", "% capacity")
  )
}

pop_base <- format(matrix_df$Baseline[matrix_df$No == 1], big.mark = ",")
pop_horiz <- format(matrix_df$Horizon[matrix_df$No == 1], big.mark = ",")
fp_delta <- matrix_df$Abs_Change[matrix_df$No == 2]
transit_horiz <- matrix_df$Horizon[matrix_df$No == 16]
transit_delta <- matrix_df$Abs_Change[matrix_df$No == 16]
hazard_delta <- matrix_df$Abs_Change[matrix_df$No == 19]
```

# 1. Executive Brief

This technical brief summarizes the spatial scenario evaluation conducted for **`r params$project_name`**. The comparative simulation benchmarks the **`r params$base_year` Baseline** against the **`r params$horizon_year` Horizon Scenario**, capturing demographic growth, land consumption, transit accessibility, and environmental hazard vulnerabilities.

Between `r params$base_year` and `r params$horizon_year`, the municipal population is projected to expand from **`r pop_base`** to **`r pop_horiz` inhabitants**. Under the simulated intervention, built-up urban land increases by **`r fp_delta` km$^2$**, while the proportion of inhabitants with direct access to formal public transit expands by **`r transit_delta` percentage points** to reach **`r transit_horiz`\\%**. Furthermore, targeted risk mitigation reduces population exposure to environmental hazards by **`r abs(hazard_delta)` percentage points**.

```{r fig-kpi-summary, fig.width=6.5, fig.height=2.4}
# Standalone Vector Graphics Chart for Illustrator/Inkscape
kpi_plot_df <- data.frame(
  Metric = factor(c("Population Density", "Transit Catchment", "Active Mobility", "Hazard Mitigation"),
                  levels = c("Hazard Mitigation", "Active Mobility", "Transit Catchment", "Population Density")),
  DeltaPct = c(-23.4, 68.2, 22.3, 5.3),
  Type = c("Negative", "Positive", "Positive", "Positive")
)

ggplot(kpi_plot_df, aes(x = DeltaPct, y = Metric, fill = Type)) +
  geom_col(width = 0.55) +
  geom_vline(xintercept = 0, color = "#556270", linewidth = 0.8) +
  scale_fill_manual(values = c("Positive" = "#4ECDC4", "Negative" = "#FF6B6B")) +
  theme_minimal(base_family = "Helvetica") +
  labs(
    title = "Macro KPI Strategic Delta (2025 vs. 2030)",
    subtitle = "Relative percentage shift across structural development pillars",
    x = "Relative Change (%)",
    y = NULL
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(color = "#556270", face = "bold", size = 12),
    plot.subtitle = element_text(color = "#88929A", size = 9),
    axis.text = element_text(color = "#556270", size = 9),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
  )
```

# 2. The Baseline Reality (`r params$base_year`)

The baseline spatial diagnostic reveals a monocentric urban structure characterized by an urban footprint of **`r matrix_df$Baseline[matrix_df$No == 2]` km$^2$** and an average density of **`r format(matrix_df$Baseline[matrix_df$No == 3], big.mark = ",")` inhabitants per km$^2$**. While the historical core maintains moderate street connectivity, peripheral extensions exhibit significant infrastructure deficits.

Currently, **`r 100 - matrix_df$Baseline[matrix_df$No == 16]`\\% of inhabitants** lack immediate proximity to formal mass transit, and **`r matrix_df$Baseline[matrix_df$No == 19]`\\% of residents** reside within designated hazard-prone zones.

# 3. The Horizon Scenario (`r params$horizon_year`)

The Horizon Scenario incorporates strategic infill corridors, dedicated bicycle tracks, and expanded transit lines. Infill policies capture **`r matrix_df$Horizon[matrix_df$No == 5]` km$^2$** of new growth within the existing urban fabric, curbing peripheral sprawl into agricultural and biodiversity conservation zones.

```{r fig-horizon-comparison, fig.width=6.5, fig.height=3.2}
# Vector figure comparing Sector Shifts
sector_df <- matrix_df %>%
  filter(No %in% c(1, 2, 9, 14, 16, 17, 19, 24)) %>%
  select(Indicator, Baseline, Horizon) %>%
  tidyr::pivot_longer(cols = c(Baseline, Horizon), names_to = "Scenario", values_to = "Value")

ggplot(sector_df, aes(x = Indicator, y = Value, fill = Scenario)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  scale_fill_manual(values = c("Baseline" = "#556270", "Horizon" = "#4ECDC4")) +
  theme_minimal(base_family = "Helvetica") +
  coord_flip() +
  labs(
    title = "Sector Performance Comparison",
    subtitle = "Baseline (Mighty Slate #556270) vs. Horizon Scenario (Pacifica #4ECDC4)",
    x = NULL,
    y = "Indicator Metric Value"
  ) +
  theme(
    plot.title = element_text(color = "#556270", face = "bold", size = 11),
    plot.subtitle = element_text(color = "#88929A", size = 8.5),
    axis.text = element_text(color = "#556270", size = 8.5),
    legend.position = "top",
    legend.title = element_blank()
  )
```

# 4. Sector Deep Dives

## Active Mobility & Public Transit
Cycling infrastructure density increases by **`r matrix_df$Pct_Change[matrix_df$No == 14]`\\%**, expanding the cycle network to **`r matrix_df$Horizon[matrix_df$No == 14]` km/km$^2$**. Consequently, population accessibility to dedicated cycle tracks jumps from **`r matrix_df$Baseline[matrix_df$No == 15]`\\%** to **`r matrix_df$Horizon[matrix_df$No == 15]`\\%**.

## Spatial Equity & Housing
The housing vacancy rate declines from **`r matrix_df$Baseline[matrix_df$No == 23]`\\%** to **`r matrix_df$Horizon[matrix_df$No == 23]`\\%**, signaling healthier occupancy. The Walkability Index improves from **`r matrix_df$Baseline[matrix_df$No == 24]`** to **`r matrix_df$Horizon[matrix_df$No == 24]`**.

## Climate Resilience & Carbon Sequestration
Urban tree plantings and conservation buffers expand annual CO2 sequestration to **`r format(matrix_df$Horizon[matrix_df$No == 20], big.mark = ",")` tonnes per year**, while Urban Heat Island intensity diminishes by **`r abs(matrix_df$Abs_Change[matrix_df$No == 26])`$^\circ$C**.

\newpage
# 5. Technical Appendix: Complete 27-Indicator Matrix

The following table presents the complete evaluation matrix of all 27 indicators formatted according to the *Libro Blanco* publishing standard.

```{r table-matrix-appendix, results='asis'}
table_subset <- matrix_df %>%
  select(No, Indicator, Baseline, Horizon, Abs_Change, Pct_Change, Units)

table_subset$Units <- gsub("%", "\\\\%", table_subset$Units)
table_subset$Units <- gsub("²", "$^2$", table_subset$Units)
table_subset$Units <- gsub("°C", "$^\\circ$C", table_subset$Units)
table_subset$Indicator <- gsub("&", "\\\\&", table_subset$Indicator)

knitr::kable(
  table_subset,
  format = "latex",
  booktabs = TRUE,
  col.names = c("No.", "Indicator", "Base", "Horizon", "$\\Delta$ Abs", "$\\Delta$ \\%", "Units"),
  escape = FALSE,
  caption = "Master 27-Indicator Urban Performance Evaluation Matrix"
)
```
