---
title: "Introduction to the *urbanperformance* Package"
output: 
  rmarkdown::pdf_document:
    extra_dependencies: ["float"]
    toc: true           
    number_sections: true
vignette: >
  %\VignetteIndexEntry{urbanperformance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# **Introduction**

This vignette introduces the `urbanperformance` package, developed within the Urban Performance (UP) Tool framework. The UP Tool is designed to generate scenarios to project possible futures for specific study areas. The tool includes two main components: the Urban Performance Calculator and the Urban Performance Visualizer.

The `urbanperformance` package integrates a list of indicators aligned with the Sustainable Development Goals (SDGs). Included in the Urban Performance Calculator, these indicators assess a city's present and future performance by modeling the impacts of investment projects and policy interventions.

Through this package, users can access functions to evaluate multiple scenarios and spatial solutions, enabling rigorous testing of their effectiveness in advancing sustainable urban development.

## Indicators

The `urbanperformance` package includes a list of indicators designated to assess city conditions across various scenarios. These indicators are numeric values that characterize the urban environment and identify critical issues, facilitating the evaluation, monitoring, and communication of a city's status. They are key for integrated urban planning, enabling users to model how an area currently addresses specific urban challenges.

The indicators included in the package are listed in the table below. Note that several of these metrics comprise sub-indicators to enhance the granularity of the analysis.

| List of indicators |  |
| :---- | :---- |
| 1. Total population | 15. Cycle proximity |
| 2. Urban footprint | 16. Public transport proximity |
| 3. Population density | 17. Amenities proximity |
| 4. Land consumption | 18. Jobs proximity |
| 5. Infill | 19. Hazard exposure |
| 6. Biodiversity land consumption | 20. Tree canopy & carbon sequestration |
| 7. Agricultural land consumption | 21. Housing costs |
| 8. Green land consumption | 22. Anthropogenic hazard exposure |
| 9. Green land per capita | 23. Vacant housing rate |
| 10. Land cover loss | 24. Walkability index |
| 11. Roads length | 25. Housing affordability ratio |
| 12. Roads density | 26. Urban heat island intensity |
| 13. Intersection density | 27. Transit utilization rate |
| 14. Cycle track density |  |

Similar to the Urban Performance Calculator, the `urbanperformance` package operationalizes the calculation methodology for these indicators. These methodologies are demonstrated in this document through a practical exercise. In essence, `urbanperformance` executes a series of geoprocessing tasks and calculations.

## Basic Data

To calculate the indicators, the `urbanperformance` package requires two types of inputs: spatial data and numerical data.

### Spatial Data

This comprises a set of georeferenced information layers, including:

| Data | Description |
| :---- | :---- |
| Population | Spatial distribution of inhabitants |
| Urban Amenities location | Location of schools, hospitals, and basic services |
| Employment distribution | Spatial distribution of employment |
| Roads network | Location of roads, including categories |
| Land cover zones | Land cover areas and classification |
| Cycle tracks | Location of cycle tracks |
| Transport systems | Location of bus stops or transport routes |
| Hazard-prone areas | Location of hazard-prone areas in the city |
| Biodiversity areas | Delimitation of biodiversity zones |
| Green areas | Location of green areas in the city |

The package supports various standard formats, including rasters (`.tif`) or vectors (`.shp`, `.kml`, `.geojson`).

*Note: All spatial datasets must be cleaned and standardized prior to use within the package.*

### Numerical Data

In addition to spatial layers, numerical inputs are required. These include proximity parameters, land use classifications, and scalar values representing specific facts or assumptions about the urban environment.

To demonstrate the functionality of `urbanperformance`, this vignette utilizes pre-loaded dummy data based on a localized area in Cancún, Mexico. Users can use this dataset to explore the indicators and assess conditions for the example city.

# The `urbanperformance` package

The exercise is integrated by four main steps to use the `urbanperformance` package:

1. Installation
2. Input data
3. Indicators calculation
4. Results analysis

## Installation

The first step is to load the necessary packages to perform the complete example. Additionally, as can be consulted in the `urbanperformance` package documentation, the calculations included use functions from other packages to simplify processes.

```{r setup, warning= FALSE, echo = TRUE, message=FALSE}
# Load require libraries
library(urbanperformance)
library(raster)
library(sf)
library(dplyr)
```

## Input data

The second step is to load the city data. This data could be gathered from open data sources or local sources, depending on availability. For this exercise, a dummy dataset is included in the package to perform an example of application of the calculations.

The example dataset is already cleaned and standardized following a specific structure, described in the `urbanperformance` documentation. This dataset includes information about population, roads, land cover, cycle infrastructure, amenities location, hazard areas, and jobs proximity.

Dataset can be loaded as follows:

```{r}
# Load spatial data
data(aoi.cun)
data(roads.cun)
data(transport.cun)
data(cycle.cun)
data(amenities.cun)
data(land.cover.classes)

## Raster data
population_base <- system.file("extdata", "POP_2025.tif",
  package = "urbanperformance"
)
population_base <- raster::raster(population_base)

population_horizon <- system.file("extdata", "POP_2030.tif",
  package = "urbanperformance"
)
population_horizon <- raster::raster(population_horizon)

buildup_base <- system.file("extdata", "Build_up_2025.tif",
  package = "urbanperformance"
)
buildup_base <- raster::raster(buildup_base)

buildup_horizon <- system.file("extdata", "Build_up_2030.tif",
  package = "urbanperformance"
)
buildup_horizon <- raster::raster(buildup_horizon)

jobs_distribution <- system.file("extdata", "Jobs.tif",
  package = "urbanperformance"
)
jobs_distribution <- raster::raster(jobs_distribution)

hazard_zones <- system.file("extdata", "Hazard.tif",
  package = "urbanperformance"
)
hazard_zones <- raster::raster(hazard_zones)

land_cover_18 <- system.file("extdata", "Land_cover_2018.tif",
  package = "urbanperformance"
)
land_cover_18 <- raster::raster(land_cover_18)

land_cover <- system.file("extdata", "Land_cover.tif",
  package = "urbanperformance"
)
land_cover <- raster::raster(land_cover)
```

The `rasterchecker()` function ensures that all spatial rasters share the same resolution, projection, and bounding box as the base population raster.

```{r, fig.width= 7, fig.height= 8}
## Verify urban extents
hazard_zones <- rasterchecker(hazard_zones, base = population_base)
buildup_base <- rasterchecker(buildup_base, base = population_base)
buildup_horizon <- rasterchecker(buildup_horizon, base = population_base)
jobs_distribution <- rasterchecker(jobs_distribution, base = population_base)
land_cover <- rasterchecker(land_cover, base = population_base)
```

## Indicators calculation

### Total population

This function calculates the total population in the urban area by adding the population located in each pixel.

This metric serves as the fundamental baseline for the entire analysis. Accurate population counts are essential for normalizing other indicators (creating per-capita metrics) and for dimensioning the scale of urban needs, such as utility demand and service coverage.

```{r, fig.width= 4, fig.height= 5, fig.pos="H", fig.align = "center"}
## Library for plotting rasters
library(rasterVis)

levelplot(population_base,
  margin = FALSE, colorkey = list(space = "bottom"),
  main = "Population distribution"
)
```

Calculate total population:

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
# library for plotting tables
library(kableExtra)

total_population <- tot_pop(population_base)

kable(total_population, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Urban Footprint

This function quantifies the total physical area occupied by artificial surfaces, built-up areas, and urban structures.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
urban_footprint <- urban_footprint(buildup_base)

kable(urban_footprint, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

```{r, fig.width= 4, fig.height= 5, fig.pos="H", fig.align = "center"}
levelplot(buildup_base,
  margin = FALSE, colorkey = list(space = "bottom"),
  main = "Urban area"
)
```

### Population density

Population density is the amount of inhabitants per built-up area. This indicator measures the intensity of land use by dividing the total population by the urban footprint area.

Density is a key driver of urban vitality and sustainability. Higher densities can support better public transport viability and local businesses, while very low densities may indicate inefficient land use. This indicator helps assess whether the city is meeting density targets required to support sustainable infrastructure.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
pop_dens <- population_density(population_base, buildup_base)

kable(pop_dens, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Land cover areas

This function calculates areas for each land cover in a specific scenario.

```{r, fig.width= 4, fig.height= 5, fig.pos="H", fig.align = "center"}
levelplot(land_cover_18,
  margin = FALSE, colorkey = list(space = "bottom"),
  main = "Land cover 2018"
)
levelplot(land_cover,
  margin = FALSE, colorkey = list(space = "bottom"),
  main = "Land cover 2025"
)
```

```{r fig.height=6, fig.pos="H", fig.width=5, message=FALSE, warning=FALSE}
land_l <- stack(land_cover_18, land_cover)
lc_areas <- land_cover_areas(land_l, land.cover.classes)

kable(lc_areas, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Land cover loss

This function calculates the amount of a specific land coverage that is become to a different land cover, between different scenarios. This indicator can be calculate for a specific land coverage loss or different coverages including in the same raster.

```{r,fig.width= 5, fig.height= 6, fig.pos="H"}
land_loss <- land_cover_loss(land_l, areas = lc_areas)

kable(land_loss, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Green land per capita

This function calculates the total green area within the urban area (forests, parks, gardens, etc.) per inhabitant.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
green_land <- land_cover[[1]]
green_land[green_land != 8] <- NA

green_a_capita <- green_area_pcapita(green_land, pop = population_base)

kable(green_a_capita, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Roads length

This function calculates the roads kilometers per each fclass in the study area.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
roads_l <- roads_length(roads.cun)

kable(roads_l, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Roads density

This function calculates the roads kilometers per square kilometer in the study area. Road density indicates the level of infrastructure penetration. While high density implies good connectivity, an excessively high density may suggest an over-dominance of car-oriented infrastructure and high impervious surface coverage, which can contribute to heat island effects and runoff issues.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
roads_dens <- roads_density(roads.cun, buildup_base)

kable(roads_dens, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Intersection density

This function calculates the average number of urban crossroads per pixel. Intersection density is a widely recognized proxy for walkability. A higher density typically indicates a finer street grain (smaller blocks) and greater connectivity, which offers pedestrians more route choices and shorter distances between destinations. Conversely, low intersection density often characterizes car-dependent suburban designs.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
intersection_dens <- intersection_density(roads.cun, buildup_base)

kable(intersection_dens, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Cycle track density

This function calculates the cycle track kilometers per square kilometer of urban area.

This measures the supply of active mobility infrastructure relative to the city size. It helps identify if the investment in cycling networks matches the scale of the urban area.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
cycle_dens <- cycle_track_density(cycle.cun, buildup_base)

kable(cycle_dens, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Cycle proximity

This function calculates the total and the percentage of the population that lives within a maximum recommended distance from a cycle track. This indicator reveals who actually has access to the network, highlighting gaps where neighborhoods are disconnected from safe cycling routes.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
cycle_prox <- cycle_proximity(cycle.cun, population_base)

kable(cycle_prox, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Public transport proximity

This function calculates the total and the percentage of the population that lives within a maximum recommended distance from a public transport station, according to the transport classification.

High proximity ensures that mass transit is a viable option for residents, reducing dependency on private vehicles and lowering carbon emissions.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
ptransport_prox <- public_transport_proximity(transport.cun,
  pop = population_base
)

kable(ptransport_prox, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Amenities proximity

This function calculates the total and the percentage of the population that lives within the maximum recommended distance to each of the following urban public services and amenities. Setting save = FALSE prevents the function from writing intermediate shapefiles to your local working directory.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
amenities_prox <- amenities_proximity(amenities.cun,
  pop = population_base,
  save = FALSE
)

kable(amenities_prox, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Jobs proximity

This function calculates the percentage of the population that lives within the areas with high job density in the urban area. This indicator helps in evaluation of the proportion of residents living near major employment clusters or high job-density zones.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
jobs_prox <- jobs_proximity(jobs_distribution[[1]], pop = population_base)

kable(jobs_prox, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

```{r, fig.width= 4, fig.height= 5, fig.pos="H", fig.align = "center"}
levelplot(jobs_distribution,
  margin = FALSE, colorkey = list(space = "bottom"),
  main = "Jobs concentration"
)
```

### Population in risk

This function calculates the number and percentage of the population that is exposed to hazards because they live within the zone of influence of natural and anthropogenic sources.

It quantifies the human risk associated with current settlement patterns and helps prioritize areas for mitigation strategies or relocation policies.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
pop_in_hazard <- hazard_exposure(hazard_zones, pop = population_base)

kable(pop_in_hazard, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

```{r, fig.width= 4, fig.height= 5, fig.pos="H", fig.align = "center"}
levelplot(hazard_zones, margin = FALSE, colorkey = list(space = "bottom"))
```

### Land consumption

This function calculates land consumption between two urban growth scenarios. Land consumption is the amount of land predicted to change from natural habitats or agricultural uses into urban human settlements between the base year and the horizon year.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
l_consumption <- land_consumption(buildup_base, buildup_horizon)

kable(l_consumption, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

*Note: Since the package avoids saving objects to the global environment to comply with CRAN policies, you can calculate and plot the urban expansion directly by computing the difference between the base and horizon footprint rasters:*

```{r, fig.width= 4, fig.height= 5, fig.pos="H", fig.align = "center"}
# Calculate urban expansion
urban_expansion <- buildup_horizon - buildup_base
urban_expansion[urban_expansion != 1] <- NA

levelplot(urban_expansion,
  margin = FALSE,
  colorkey = list(space = "bottom")
)
```

### Infill

This function calculates the amount of land that was located within the boundaries of the urban area in the base scenario and that at least doubled its population between the base year and the horizon year. It is also known as intensification area.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
infill_area <- infill(population_base, population_horizon)

kable(infill_area, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Tree Canopy Cover and Carbon Sequestration

This function projects tree population growth, public canopy per capita, and metric tonnes of CO2 sequestered using regional species parameters and planting targets.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
# Forestry parameters for regional native species in Cancún
tree_params_df <- data.frame(
  species = c("Chacáh", "Jabín", "Chechén", "Ceiba"),
  canopy_sqm = c(50.2, 60.0, 45.0, 175.0),
  foliar_co2seq_kg_yr = c(22.5, 25.5, 21.0, 45.0),
  annual_mortality_rate = c(0.025, 0.020, 0.018, 0.015),
  years_to_maturity = c(8, 12, 14, 15)
)

tree_canopy_res <- tree_canopy_carbon(
  trees_base = NULL,
  pop = population_base,
  params = tree_params_df,
  horizon_years = 10,
  annual_forestation = 3500,
  public_space_ratio = 0.45
)

kable(tree_canopy_res, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Housing Costs

This function estimates the average housing unit cost based on land value, construction cost, and modal socioeconomic deciles across residential blocks.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
# Cost multipliers across socioeconomic deciles
decile_costs <- data.frame(
  decile = 1:10,
  land_cost_sqm_usd = c(45, 60, 85, 300, 350, 400, 450, 800, 1200, 2500),
  constr_cost_sqm_usd = c(324, 380, 450, 810, 900, 980, 1135, 1350, 1800, 2432),
  avg_unit_sqm = c(42, 45, 55, 68, 80, 95, 120, 180, 220, 350)
)

# Simulate residential blocks from AOI
res_blocks <- suppressWarnings(st_buffer(st_sample(aoi.cun, size = 20), dist = 150))
res_blocks <- st_sf(id = 1:20, geometry = res_blocks)

# Decile raster
decile_r <- raster(extent(population_base), res = res(population_base), crs = crs(population_base))
set.seed(42)
values(decile_r) <- sample(1:10, ncell(decile_r), replace = TRUE)

hc_res <- housing_costs(res_blocks, decile_r, decile_costs)

kable(hc_res, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Anthropogenic Hazard Exposure

This function calculates population exposed to localized pollution and industrial hazards using outreach buffers with spatial unioning to eliminate double-counting.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
# Sample pollution sources in Cancún
poll_pts <- st_sf(
  fclass = c("water_contamination", "industrial_stationary"),
  geometry = st_sfc(
    st_point(c(516000, 2335000)),
    st_point(c(518000, 2338000))
  ),
  crs = st_crs(aoi.cun)
)

hazard_exp <- anthropogenic_hazard(poll_pts, population_base)

kable(hazard_exp, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Vacant Housing Rate

This function calculates the percentage of vacant units across the city to detect speculative vacancies or housing market failures.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
# Total housing units based on population raster
hu_total <- population_base / 3.5
hu_vacant <- hu_total * 0.062 # 6.2% simulated frictional/speculative vacancy

vac_res <- vacant_housing(hu_total, hu_vacant)

kable(vac_res, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Walkability Index

This function evaluates pedestrian accessibility by combining intersection connectivity, transit access, and destination diversity across a 500m analysis grid.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
walk_res <- walkability_index(
  roads = roads.cun,
  transit_stops = st_cast(st_sample(transport.cun, 30), "POINT"),
  footprint = buildup_base,
  amenities = amenities.cun,
  grid_size_m = 500,
  weights = c(0.33, 0.33, 0.34)
)

kable(walk_res, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Housing Affordability Ratio

This function computes the Price-to-Income Ratio (PIR) and classifies housing markets according to UN-Habitat and World Bank standards.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
afford_res <- housing_affordability(
  housing_cost = hc_res$value[1],
  income = 972, # Median monthly income (Decile 5)
  income_period = "monthly"
)

kable(afford_res, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Urban Heat Island (UHI) Intensity

This function quantifies the thermal differential between urban and natural rural surroundings, strictly excluding water bodies from rural baseline estimation.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
# Simulated Landsat 8 LST raster
lst_sim <- raster(extent(land_cover), res = res(land_cover), crs = crs(land_cover))
set.seed(123)
# Base temperature around 27C, urban pixels elevated to 32C
values(lst_sim) <- 27.5 + rnorm(ncell(lst_sim), sd = 0.8)
lst_sim[land_cover == 2] <- lst_sim[land_cover == 2] + 4.5

uhi_res <- uhi_intensity(lst_sim, land_cover, rural_buffer_km = 3, water_class = 3, urban_class = 2)

kable(uhi_res, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Transit Utilization Rate

This function assesses mobility system capacity versus daily demand, tagging routes against municipal transit benchmarks.

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
cancun_transit <- data.frame(
  transit_mode = c("trunk_brt", "standard_bus"),
  nominal_vehicle_capacity = c(100, 45),
  crush_vehicle_capacity = c(150, 70),
  daily_trips = c(120, 80),
  daily_riders = c(9800, 3200)
)

transit_res <- transit_utilization(cancun_transit, capacity_type = "both")

kable(transit_res, "latex", booktabs = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Results analysis

Select proximity data for visualization

```{r, fig.width= 5, fig.height= 6, fig.pos="H"}
## Join all proximities in a data frame
proximities <- rbind(cycle_prox, ptransport_prox, amenities_prox, jobs_prox)

## Select only percentages
proximities_pct <- subset(proximities, units == "%")
```

Visualize results

```{r, fig.width= 5, fig.height= 4, fig.pos="H", fig.align = "center"}
## library for plotting proximity results
library(ggplot2)
p <- ggplot(proximities_pct, aes(x = fclass, y = value, fill = fclass)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set2") +
  theme_bw() +
  ylab("Percentage of inhabitants with proximity") +
  xlab("Urban amenities and services") +
  theme(legend.position = "none") +
  ylim(0, 100)

p
```
